Revisiting the hydrodynamic modulation of short waves by longer waves

### 2024-09-19 paper/paper.tex
Plotting inhomogeneity tendencies

\begin{abstract}
Hydrodynamic modulation of short ocean surface waves by longer ambient waves is
a well-known ocean surface process that affects remote sensing, the
interpretation of \textit{in situ} wave measurements, and numerical wave
forecasting.
In this paper, we revisit the linear wave theory and derive higher-order
solutions for the change of short-wave wavenumber, action density, and
gravitational acceleration due to the presence of longer waves.
We validate the analytical solutions with numerical simulations of the full wave
crest and action conservation equations.
The nonlinear analytical solutions of short-wave wavenumber, amplitude, and
steepness modulation significantly deviate from the linear analytical solutions
of \citet{longuet1960changes}, and are similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution}.
We also examine the result of \citet{peureux2021unsteady} who found through
numerical simulations that the short-wave modulation grows unsteadily with
each long-wave phase.
We show that this unsteady growth only occurs for homogeneous initial
conditions and does not occur for more realistic long-wave groups.
The combined analytical and numerical framework presented here allows us to
quantify the importance of various terms in the full wave conservation
equations.
The new analytical solutions are a good approximation of the full numerical
solution in long-wave steepness up to $\approx 0.2$.
\end{abstract}

@@

To quantify the contribution of the nonlinear terms in Eqs. \ref{eq:wavenumber}
and \ref{eq:wave_action2}, and to evaluate the steadiness assumption of the
analytical solutions, we now proceed to numerically integrate the full set of
crest and action conservation equations.
Equations (\ref{eq:wavenumber})-(\ref{eq:wave_action2}) are discretized using
second-order central finite difference in space and integrated in time using the
forth-order Runge-Kutta method \citep{butcher1996runge}.
The space is divided into 100 grid points.
This is effectively the same equation set and numerical configuration as that of
\citet{peureux2021unsteady}, except that their long-wave orbital velocities are
evaluated at $z=0$ instead of $z=\eta_L$, which neglects the Stokes drift of
short-wave groups.
Their gravity modulation also does not consider the nonlinear or the curvature
effects, however, this difference does not qualitatively affect the results.
Another difference is the choice of the numerical scheme for spatial differences,
which in their case is the MUSCL4 scheme by \citet{kurganov2000new}.
Although the second-order central finite difference is not appropriate for many
numerical problems, in our case it is sufficient because it is conservative and
the fields that we compute derivatives of are smooth and continuous.
The maximum relative error of the centered finite difference in this model
is $\approx 0.066\%$.
We show in the next section that our numerical solutions for infinite long-wave
trains are qualitatively equivalent to those of \citet{peureux2021unsteady}.

@@

The key result of \citet{peureux2021unsteady} is that of unsteady steepening of
short waves when a homogeneous field of short waves is forced by an infinite
long-wave train.
The steepening is driven by the progressive increase in the action of short
waves (as opposed to their wavenumber), and is centered on the group of short
waves that begins its journey in the trough of the long wave.
This is because the unmodulated short waves that begin in the trough enter into
the area where the surface velocities are converging.
Inversely, the short waves that begin around the crest of the long-wave will
enter into the area of surface divergence and will thus have their wave action
and wavenumber, decreased.
We reproduce this behavior in Fig. \ref{fig:modulation_3panel_infinite}, which
shows the evolution of the short-wave action, wavenumber, and steepness
modulation (here defined as the change of a quantity relative to its initial value).
After 10 long-wave periods, the short-wave action modulation $\widetilde{N}/N$
is approximately doubled, which almost exactly corresponds to the unsteady growth
reported by \citet{peureux2021unsteady} in their Fig. 1.
Another important property of this solution is that the short-wave group that
experiences peak amplification (dampening) are not locked-in to the long-wave
crests (troughs), as they are in the L-HS solution.

@@

Why does this unsteady growth of short-wave action occur?
As \citet{peureux2021unsteady} explain, "the additional change in advection
velocity due to the modulation of the short wave group speed gives an additional
amplification."
That is, it is the inhomogeneity term $N \dfrac{\partial C_g}{\partial x}$ that
is responsible for the instability.
However, the instability only occurs if the short waves are initialized as a
uniform field.
If they are initialized instead as a periodic function of the long-wave phase
such that their wavenumber (and optionally, action) is higher on the crests
and lower in the troughs, akin to the L-HS solution, the instability does not
occur.

@@

Why is it, then, that the solution is so different depending on the short-wave
initial conditions?
\citet{peureux2021unsteady} did not provide an explanation for this behavior.
At first, it may seem that there is something special about the uniform initial
conditions that makes its solutions grow unsteadily and indefinitely.
\citet{peureux2021unsteady} describe this scenario as "...the sudden appearance
of a long wave perturbation in the middle of a homogeneous short wave field".
The sudden appearance of long waves may be a clue.
The exact numerical simulation as shown in Fig. \ref{fig:modulation_3panel_infinite},
except for a linear ramp applied to $a_L$ (and consequently, $U$ and $W$ as well)
for the first five periods reveals that the unsteady growth no longer occurs
(Fig. \ref{fig:modulation_3panel_ramp}).

@@

@@

>>>>

\begin{abstract}
Hydrodynamic modulation of short ocean surface waves by longer ambient waves is
a well-known ocean surface process that affects remote sensing, the
interpretation of \textit{in situ} wave measurements, and numerical wave
forecasting.
In this paper, we revisit the linear wave theory and derive higher-order
solutions for the change of short-wave wavenumber, action density, and
gravitational acceleration due to the presence of longer waves.
We validate the analytical solutions with numerical simulations of the full wave
crest and action conservation equations.
The nonlinear analytical solutions of short-wave wavenumber, amplitude, and
steepness modulation significantly deviate from the linear analytical solutions
of \citet{longuet1960changes}, and are similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution}.
We also examine the result of \citet{peureux2021unsteady} who found through
numerical simulations that the short-wave modulation grows unsteadily with
each long-wave phase.
We show that this unsteady growth only occurs for homogeneous initial
conditions and does not occur for more realistic long-wave groups.
The combined analytical and numerical framework presented here allows
quantifying the importance of various terms in the full wave conservation
equations.
The proposed analytical solutions are a good approximation of the fully
nonlinear numerical solutions in long-wave steepness up to $\approx 0.2$.
\end{abstract}

@@

To quantify the contribution of the nonlinear terms in Eqs. \ref{eq:wavenumber}
and \ref{eq:wave_action2}, and to evaluate the steadiness assumption of the
analytical solutions, we now proceed to numerically integrate the full set of
crest and action conservation equations.
Equations (\ref{eq:wavenumber})-(\ref{eq:wave_action2}) are discretized using
second-order central finite difference in space and integrated in time using the
forth-order Runge-Kutta method \citep{butcher1996runge}.
The space is divided into 100 grid points.
This is effectively the same equation set and numerical configuration as that of
\citet{peureux2021unsteady}, except that their long-wave orbital velocities are
evaluated at $z=0$ instead of $z=\eta_L$, thus neglecting the Stokes drift
induced by long waves on short-wave groups
\citep{stokes1846,van2018stokes,monismith2020stokes}.
Their gravity modulation also does not consider the nonlinear or the curvature
effects, however, this difference does not qualitatively affect the results.
Another difference is the choice of the numerical scheme for spatial differences,
which in their case is the MUSCL4 scheme by \citet{kurganov2000new}.
Although the second-order central finite difference is not appropriate for many
numerical problems, in our case it is sufficient because it is conservative and
the fields that we compute derivatives of are smooth and continuous.
The maximum relative error of the centered finite difference in this model
is $\approx 0.066\%$.
We show in the next section that our numerical solutions for infinite long-wave
trains are qualitatively equivalent to those of \citet{peureux2021unsteady}.

@@

The key result of \citet{peureux2021unsteady} is that of unsteady steepening of
short waves when a homogeneous field of short waves is forced by an infinite
long-wave train.
The steepening is driven by the progressive increase in the action of short
waves (as opposed to their wavenumber), and is centered on the group of short
waves that begins its journey in the trough of the long wave.
This is because the unmodulated short waves that begin their journey in the
trough first enter the convergence region before reaching the divergence region.
Inversely, the short waves that begin around the crest of the long-wave first
enter the divergence region and will thus have their action and wavenumber
decreased.
We reproduce this behavior in Fig. \ref{fig:modulation_3panel_infinite}, which
shows the evolution of the short-wave action, wavenumber, and steepness
modulation (here defined as the change of a quantity relative to its initial value).
The long waves have $\varepsilon_L = 0.1$ and $k_L = 1$, and the short waves
have $k = 10$.
After 10 long-wave periods, the short-wave action modulation $\widetilde{N}/N$
is approximately doubled, consistent with the unsteady growth reported by
\citet{peureux2021unsteady}.
Another important property of this solution is that the short-wave group that
experiences peak amplification (dampening) are not locked in to the long-wave
crests (troughs), as they are in the solution by \citet{longuet1960changes}.

@@

Why does this unsteady growth of short-wave action occur?
As \citet{peureux2021unsteady} explain, "the additional change in advection
velocity due to the modulation of the short wave group speed gives an additional
amplification."
That is, it is the inhomogeneity term $N \dfrac{\partial C_g}{\partial x}$ that
is responsible for the instability.
However, the instability only occurs if the short waves are initialized as a
uniform field.
If they are initialized instead as a periodic function of the long-wave phase
such that their wavenumber (and optionally, action) is higher on the crests
and lower in the troughs, akin to the prior steady solutions, the instability
does not occur.

@@

Why is it, then, that the solution for short-wave modulation is not only
quantitatively, but also qualitatively, different depending on the short-wave
initial conditions?
It may at first seem that there is something special about the uniform initial
conditions that makes its solutions grow unsteadily and indefinitely.
\citet{peureux2021unsteady} described this scenario as "...the sudden appearance
of a long wave perturbation in the middle of a homogeneous short wave field".
However, such a scenario is not realistic in the ocean where waves tend to
organize into groups due to their dispersive nature.
To test whether it is the sudden onset of long-wave forcing that causes the
unstable growth, we repeat the numerical simulation of Fig.
\ref{fig:modulation_3panel_infinite}, except for a linear ramp applied to $a_L$
(and consequently, $U$ and $W$ as well) for the first five periods:
$a_L(t) = min(a_L, a_L t / \left(5 T_L\right))$, where $T_L$ is the long-wave period.
This result is shown in Fig. \ref{fig:modulation_3panel_ramp}.
The key distinction that arises with the introduction of the ramp is that the
modulation pattern of short waves is now locked in with the long-wave crests,
and their indefinite steepening vanishes.
The stabilization effect of the long-wave ramp is more readily assessed in
Fig. \ref{fig:unsteady_growth_timeseries}, which shows the maximum short-wave
steepness modulation as function of time over 30 long-wave periods, for the
infinite long-wave train case and that when applying the linear ramp to the
long-wave forcing.
Comparing the wave action propagation ($C_g \frac{\partial N}{\partial x}$) and
inhomogeneity ($N \frac{\partial C_g}{\partial x}$) tendencies reveals that
after only one long-wave period, the propagation and inhomogeneity tendencies
are unbalanced and amplify the steepening of short waves
(Fig. \ref{fig:inhomogeneity_tendencies}).
Although \citet{peureux2021unsteady} correctly pointed out that the unstable
growth of short-wave steepness occurs due to the resonance of the inhomogeneity
tendency with the short-wave action perturbation, in case of a gentle ramp
up of the long-wave forcing, the propagation and inhomogeneity tendencies
mostly balance each other out, thus allowing the short-wave steepening to
lock in to the long-wave crests.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{figures/inhomogeneity_tendencies.png}
  \caption{
    Propagation ($C_g \frac{\partial N}{\partial x}$) (blue) and inhomogeneity
    ($N \frac{\partial C_g}{\partial x}$) (orange) tendencies and their sum
    (green) for the short-wave action after one long-wave period, in the case of
    (a) infinite long-wave train and (b) linear ramp of long-wave amplitude.
    Dashed line is the long-wave elevation and the dotted line is the short-wave
    action modulation minus 1.
  }
  \label{fig:inhomogeneity_tendencies}
\end{figure}

@@

\section{Sensitivity to wavenumber ratios}
\label{section:sensitivity_to_wavenumber_ratios}
